AAnalyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2025-01-14 CSEDU2025/Paper.tex
paper update

\title{
  Analyzing Student Use of Spacing and Interleaving Strategies in Interactions
  with GenAI-Powered Chatbots in Programming Courses
}

@@

\cite{Bastani24} investigated the impact of GenAI, specifically GPT-4, on human
learning, with a focus on mathematics education in a high school. The problem
addressed was how the use of generative AI could affect the acquisition of new
skills, which is crucial for long-term productivity. The method involved a
controlled experiment with approximately one thousand students, who were exposed
to two GPT-4-based tutors: a simple tutor (GPT Base) and another with safeguards
designed to promote learning (GPT Tutor). The results showed that while access
to GPT-4 improved performance on practice exercises (48\% with GPT Base and
127\% with GPT Tutor), the removal of access to GPT Base led to a 17\% decrease
in student performance on exams. This suggests that unrestricted use of
\textit{GPT Base} could hinder learning. However, the GPT Tutor was able to
mitigate this negative effect.

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{4cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change (LC) & Prompts that request a change of idiom. \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

When a student requests a code snippet (CS), it is generally predictable that
they will resume interaction with the chatbot within a short period of time, as
indicated by the mean (1.26) and standard deviation (1.74) of the Code Snippet
category in Table \ref{tab:averages}. This result is noteworthy because,
although some requests are simple and lead to quick responses, others may be
more complex and would require more time for reflection. However, the low
average response time and minimal variability may indicate that students adopt a
more passive learning posture when requesting the chatbot to generate small pieces
of code.

@@

Despite some occasional variation (3.54), prompts classified as Uncategorized
exhibit a relatively low average time (1.65) for students to resume interaction
with the chatbot. This variation occurs, in part, because of situations where students
end the conversation with an uncategorized message, such as a simple
\textit{"thanks"}, but later choose to resume the interaction.

When a student requests the chatbot to change the language (LC) in a response, the
new interaction occurs quickly (1.89) with little variation in response time
(3.28), meaning that it is predictable that the student will continue interacting.

Although still below the overall average, students take a moderate amount of
time (2.65) to resume interaction with the chatbot after requesting the solution to
a multiple-choice question (MQ). The variation (6.10) suggests that some
responses may lead to a slower follow-up interaction, possibly indicating a
moment of deeper reflection on the part of the students. However, since both the
mean and standard deviation are below the overall average, this could indicate
a more passive behavior, as students obtain ready-made answers from the chatbot.

The average time for students to resume interaction in the correction category
(SC) is close to the overall average (3.17), although the high standard
deviation (7.33) indicates significant variation. In some cases, students are
already familiar with the answer and only seek a quick correction from the chatbot;
in others, they detect flaws in the chatbot's responses and need to carefully
examine the subsequent answer to identify potential errors, which suggests a
more active behavior.

After a question classified as Debugging Help (DH), there is both a
longer-than-average response time (4.84) and a high variation in the time taken
for a new interaction (10.13). This suggests that student prompts in this
category tend to be more time-consuming and unpredictable. A possible
explanation is that some DH issues are simple, such as '\textit{what does x++ do?}',
while others involve requests for analysis of more complex code, like
\textit{'explain this piece of code (accompanied by a complete method).'}

Conceptual questions (CQ) generally result in a high average time (5.41) for
students to reengage with the chatbot. Moreover, the high standard deviation
(10.54) indicates an unpredictable variation in response times. This
unpredictability may be attributed to the fact that, in some cases, the
complexity of the question and answer demands more reflection time from the
student, whereas in others, lesser complexity allows for a quicker interaction
with the chatbot.

Finally, a message classified as a Complete Solution request (SCO) exhibits the
longest average time (6.55) for students to resume interaction with the chatbot.
Additionally, the high standard deviation (10.14) highlights significant
unpredictability in response time. After requesting a complete solution,
students tend to take longer to return due to the complexity of the problems
involved, such as the implementation of sorting algorithms such as Heapsort. Table
\ref{tab:averages} presents the means and standard deviations for each
category.

However, some categories suggest quicker interactions with less reflection on
the part of students, such as 'Code Snippet' and 'Multiple Choice.' On the other
hand, other categories indicate that students may be in a more reflective
process, such as 'Debugging Help,' 'Conceptual Question,' and 'Complete
Solution.' This demonstrates that the level of student engagement varies
according to the nature of the requested interaction, reflecting different
approaches and depths in the learning process.

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{4cm} p{4cm} p{4cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.26 & 1.74 \\
      Uncategorized & 1.65 & 3.54 \\
      Language Change & 1.89 & 3.28 \\
      Multiple Question & 2.65 & 6.10 \\
      Student Correction & 3.17 & 7.33 \\
      \hdashline
      \textit{Overall} & \textit{3.43} & \textit{6.60} \\
      \hdashline
      Debugging Help & 4.84 & 10.13 \\
      Conceptual Questions & 5.41 & 10.54 \\
      Complete Solution & 6.55 & 10.14 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

The objective of this work was to analyze the interactions of Java programming
students with an educational chatbot developed using the RAG technique, employing
two metacognitive study strategies: spacing and interleaving.

The metacognition theory, along with the classification of student actions,
proved to be effective in understanding and revealing how students use the chatbot.
The results showed that most messages fall into the categories of Conceptual
Questions and Debugging Help, both known for promoting slower engagement and
potentially fostering more active learning. However, the results also indicated
that students' use of study strategies such as spacing and interleaving remains
limited, from which we can conclude that encouraging students to apply these
strategies more intentionally could enhance their learning while using GenAI.

It is important to consider how GenAI tools impact students in various
demographic groups, academic disciplines, cultural backgrounds, and levels of
previous experience \citep{catalan21} \citep{neo22}. Consequently, this study is
limited to a specific group of students and focuses on the use of a single GenAI
tool. As a result, the findings may not be generalizable to other populations or
tools.

To further explore this, a broader study is planned to examine, for instance,
the relationship between student profiles, use of metacognitive scaffolds, their
effective study strategies in using a GenAI, and learning outcomes.
In addition, to help educators analyze student behavior and provide
more targeted assistance, an AI model is being developed to automatically
classify student messages, offering valuable information on how these tools are used.

>>>>

\title{
  AAnalyzing Student Use of Spacing and Interleaving Strategies in Interactions
  with GenAI-Powered Chatbots in Programming Courses
}

@@

\cite{Bastani24} investigated the impact of GenAI, specifically GPT-4, on human
learning, with a focus on mathematics education in a high school. The problem
addressed was how the use of generative AI could affect the acquisition of new
skills, which is crucial for long-term productivity. The method involved a
controlled experiment with approximately one thousand students, who were exposed
to two GPT-4-based tutors: a simple tutor (GPT Base) and another with safeguards
designed to promote learning (GPT Tutor). The results showed that while access
to GPT-4 improved performance on practice exercises (48\% with GPT Base and
127\% with GPT Tutor), the removal of access to GPT Base led to a 17\% decrease
in student performance on exams. This suggests that unrestricted use of
\textit{GPT Base} could hinder learning. However, the GPT Tutor was able to
mitigate this negative effect.
\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.2} % Increase the spacing between rows
    \begin{tabular}{p{3.8cm} p{6cm} p{4.5cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Example of prompt} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \textit{Would this code be ok? \{code\}} \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. & \textit{What does it mean for a method to be static?} \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. & \textit{The correct answer is B} \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. & \textit{A class inherits from another write this code} \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. & \textit{Give me the code for a selection sort} \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). & \textit{A heap is a data structure appropriate for: \{options\}} \\
      Language Change (LC) & Prompts that request a change of idiom. & \textit{In Spanish} \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. & \textit{Thanks} \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

The objective of this research question was to study the time a student takes to
re-engage with the bot after sending a message for each category. This analysis
can offer insights into students' behavior: faster interactions might reflect
less reflection, whereas longer pauses could indicate the opposite.

To calculate the average time and standard deviation for messages within a
category initially was first observed conversations in which message timing was
influenced by previous messages. Pearson's correlation analysis revealed that
28.40\% of the conversations exhibited a strong correlation, which, in this
study, was defined as values outside the range of -0.3 to 0.3. In this context,
a strong correlation suggests that the response time for a message may depend
on the time of the preceding message or multiple prior messages. Consequently,
these 28.40\% of conversations were excluded from the average time and standard
deviation calculation, leaving 56 out of the initial 86 conversations. Table
\ref{tab:averages} presents the means and standard deviations for each category.

When a student requests a Code Snippet (CS), it is generally predictable that
they will summarize interaction with the chatbot within a short period, as
indicated by the mean (1.34) and standard deviation (1.95) of the Code Snippet
category in Table \ref{tab:averages}. This result is noteworthy because,
although some requests are simple and lead to quick responses, others may be
more complex and require more reflection time. However, the low average response
time and minimal variability may indicate that students adopt a more passive
learning posture when requesting the chatbot to generate pieces of code.

@@

Despite occasional variation (3.59), prompts classified as Uncategorized exhibit
a relatively low average time (1.70) for students to resume interaction with the
chatbot. This variation occurs partly because of situations where students end
the conversation with an Uncategorized message like \textit{"thanks"} but later
decide to resume the interaction.

When a student requests the chatbot to Change the Language (LC)  in a response,
in our context English to Spanish, the new interaction occurs quickly (2.23)
with slight variation in response time (3.71), meaning that it is predictable
that the student will continue interacting.

Although still below the overall average, students take moderate time (2.47) to
resume interaction with the chatbot after requesting the solution to a Multiple
Question (MQ) exercise. The variation (5.56) suggests that some responses may
lead to a slower follow-up interaction, possibly indicating a moment of deeper
reflection on the part of the students. However, since the mean and standard
deviation are below the overall average, this could indicate a more passive
behavior, as students obtain ready-made answers from the chatbot.

The average time of the Student Correction (SC) category is close to the overall
average (4.07), although the high standard deviation (8.42) indicates
significant variation. In some cases, students know the answer to an exercise
and only accomplish a quick correction from the chatbot; in others, they detect
flaws in the chatbot's responses and need to examine the subsequent answer to
identify potential errors, which suggests a more active behavior.

A message classified as Debugging Help (DH) has both a long average response
time (5.30) and high variation (10.80). Consequently, the behavior in this
category is more time-consuming and unpredictable. A possible explanation is
that some DH messages are simple, such as '\textit{what does x++ do?}'. In
contrast, others involve requests for analysis of more complex code, like
\textit{'explain this piece of code (accompanied by a complete method).'}

Conceptual Questions (CQ) generally result in a high average time (5.47) for
students to reengage with the chatbot. Moreover, the high standard deviation
(10.46) indicates an unpredictable variation in response times. This
unpredictability may be attributed to the fact that, in some cases, the
complexity of the question and answer demands more reflection time from the
student, whereas in others, lesser complexity allows for a quicker interaction
with the chatbot.

Finally, a message classified as a Complete Solution request (SCO) exhibits the
longest average time (7.65) for students to resume interaction with the chatbot.
Additionally, the high standard deviation (11.03) highlights significant
unpredictability in response time. After requesting a complete solution,
students may take longer to return due to the complexity of the problems
involved, such as the implementation of sorting algorithms such as Heapsort.

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.2} % Adjust the spacing between rows
    \begin{tabular}{p{4cm} p{3cm} p{3cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.34 & 1.95 \\
      Uncategorized & 1.70 & 3.59 \\
      Language Change & 2.23 & 3.71 \\
      Multiple Question & 2.47 & 5.66 \\
      \hdashline
      \textit{Overall} & \textit{3.78} & \textit{6.95} \\
      \hdashline
      Student Correction & 4.07 & 8.42 \\
      Debugging Help & 5.30 & 10.80 \\
      Conceptual Questions & 5.47 & 10.46 \\
      Complete Solution & 7.65 & 11.03 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

This study aimed to analyze the interactions of Java programming students with
an educational chatbot developed using the RAG technique, employing two
metacognitive study strategies: Spacing and Interleaving.

The results indicated that messages from the Code Snippets and Multiple Question
(Quiz) category implied quick and consistent interactions with the bot. This
pattern reflects a good level of engagement with the tool but also raises the
possibility of a lack of more profound reflection on the received content, given
that students' subsequent messages were sent consistently within short intervals.

On the other hand, it was observed that most students' messages concentrated on
categories such as Debugging Help and Conceptual Questions, which might suggest
that students were committed to understanding the concepts. However, the results
also indicated that students' use of study strategies such as Spacing and
Interleaving remains limited. Therefore, encouraging students to apply these
strategies intentionally could enhance their learning using GenAI.

It is fundamental to consider how GenAI tools impact students in various
demographic groups, academic disciplines, cultural backgrounds, and levels of
previous experience \citep{catalan21} \citep{neo22}. Consequently, this study is
limited to a specific group of students and focuses on using a single GenAI
tool. As a result, the findings may not be generalizable to other populations or
tools.

There are ways to deepen the study; based on previous references, 60 minutes was
the standard spacing measure. However, further studies will deepen this analysis
using a quantitative variable to represent different spacing intervals better.
Future research can also include additional metacognitive strategies, which will
involve analyzing and interpreting user messages in depth. For instance, the
Monitoring strategy was evident when students expressed that a concept remained
unclear and requested further clarification. Another necessary analysis involves
investigating the relationship between student profiles, metacognitive
strategies for using GenAI, and learning outcomes.
